---
title: "Medartis Wrist Plates"
subtitle: "Experiment 2 - bending and shear matched"
date: 2024-07-23
format: 
  html: default
  pdf: default
  docx: default
---

## setup

Load any needed libraries and define some constants.

```{r}
#| label: setup
#| warning: false
#| message: false

library(dplyr) # for manipulating dataframes
library(ggplot2) # for plotting

# If you don't have bluer install it with 
# install.packages("devtools")
# devtools::install_github(repo = "yadbor/bluer")
library(bluer) # routines for analysing mechanical test data

data_root <- here::here("data-raw")

data_col_names <- "File Number,U [mm],V [mm],W [mm],Time_1,position,load"
data_cols <- stringr::str_split_1(data_col_names, ",")
```

## read the data files

Use the code below to read all of the .csv files output from VIC-3D.
```{r}
#| label: read_raw_files
#| 
raw_files = list.files(path = here::here("data-raw", "cadaver_csv_files"),
                       pattern = "csv$",
                       recursive = TRUE,
                       full.names = TRUE,
                       ignore.case = TRUE
                       )
raw_files <- raw_files |>
  stringr::str_subset(pattern = "(?i)Marker_Data") # Case insensitive match needed
# Name the vector with the filename, which bind_rows will turn into a column
names(raw_files) <- basename(raw_files)
# Map the function read_csv() across all elements in raw_files, then bind the resulting dataframes
# rowwise into a combined dataframe, putting the names from raw_files into a column "filename".
all_results <- raw_files |>
  purrr::map(\(f) readr::read_csv(file = f, show_col_types = FALSE)) |>
  bind_rows(.id = "filename") |>
  # drop unused columns
  select(filename, all_of(data_cols)) |>
  # clean up and simplify names
  rename(frame = "File Number", u = "U [mm]", v = "V [mm]", w = "W [mm]", time = "Time_1") |>
  # fix the spelling of parts of the filename
  # First fix a known typo
  mutate(filename = stringr::str_replace(filename, "(?i)poximal", "proximal")) |>
  # Then make the marker labels all lower case
  mutate(filename = stringr::str_replace(filename, "(?i)distal", "distal")) |>
  mutate(filename = stringr::str_replace(filename, "(?i)proximal", "proximal"))
```

```{r}
#| label: parse_filenames

all_results <- all_results |>
  tidyr::separate_wider_regex(cols = filename,
                              patterns = c(id = "^[^_]+", "_", 
                                           arm = "[^_]+", "_(?i)plate\\d?", 
                                           plate = "[L|S]", "_",
                                           marker = "(?i)distal|proximal", ".*"),
                                cols_remove = FALSE,
                                #too_few = "debug"
    ) 
```

Check that the filenames have parsed to something sensible
```{r}
#| label: check_parsed

all_results |> select(id:marker) |> purrr::map( \(c) unique(c))
```
Can then write the cleaned results out.

```{r}
#| label: write_cleaned_data
readr::write_csv(all_results, file = here::here("data", "cadaver_all_results.csv"))
```

If the data files have already been cleaned up and combined, 
just read the processed data from `cadaver_all_results.csv`.

```{r}
#| label: read_data
#| eval: false # change this to read the file
all_results <- readr::read_csv(file = here::here("data", "cadaver_all_results.csv"),
                               show_col_types = FALSE)
```

## analyse cycles

Each test has multiple load-unload cycles. 
Data are recorded on the 9^th^ and 10^th^ cycles, and this is repeated 
every 10 cycles up to a total of 60 cycles.
We are interested in the loading part of each cycle.

Divide test into cycles by finding peaks and troughs, which are defined as local
maxima and minima.
The loading phase is the one going towards a peak, and the unloading phase is
towards a trough.

The `bluer` package has routines for finding peaks and labelling cycles 
and phases.

But first, as these tests are all in compression, both the load and position 
are negative.
Invert them both so that graphs read better for most people.

```{r}
#| label: label_cycles

cycles <- all_results |> 
  mutate(position = -1.0 * position, load = -1.0 * load) |> # Invert these axes
  mutate(across(id:marker, factor)) |> # Convert all id columns to factors
  group_by(id, arm, plate, marker) |> # group each test
  mutate(as_tibble(bluer::label_cycles(position))) |> # and label the cycles
  mutate(uid = paste0(id, arm, plate, marker, cycle)) # uniquely identify each cycle
```


```{r}
#| label: check_plots
# Check what we have done

cycles |> 
  filter(seg == "load") |>
  ggplot() + 
  aes(x = time, y = position, colour = marker, group = id) +
  geom_point() +
  facet_grid(plate + marker ~ id, scales = "free") +
  labs(title = "everything, everywhere, all at once")
```
There are no results for the `MSD601` short plate test, 
and only 3 cycles in the long plate test, so drop `MSD601`.

`MSD664` short plate has an extra cycle, which looks like the load to failure 
test was accidentally included in the analysis.
Drop the extra cycle, so every test has 12 cycles (6 groups of 2 cycles).

```{r}
#| label: drop_bad_tests

cycles <- cycles |>
  filter(id != "MSD601") |>
  filter(cycle < 13)
```


Plot just the loading portion of each cycle, as load vs position.
```{r}
#| label: plot_load_position

loading <- cycles |> 
    filter(seg == "load")

loading |>
    ggplot() + 
    aes(x = time, y = position, colour = factor(cycle), group = uid) +
    geom_line() +
    facet_grid(plate + marker ~ id, scales = "free") +
    guides(colour=FALSE) +
    labs(title = "position vs time - by plate, marker and id")
```
The short plate test for MSD664 appears to have run twice as fast as the 
long plate test, which needs more analysis.

It is also obvious that there is a big time gap on some cycles, visible as a 
horizonatal or near horizontal line.
Look more closely at one test to see what is happening:
```{r}
#| label: single_test

loading |>
  filter(id == "MSD575") |>
  ggplot() + 
    aes(x = time, y = position, colour = factor(cycle), group = uid) +
    geom_line() + geom_point() +
    facet_grid(plate + marker ~ id, scales = "free") +
    guides(colour=FALSE) +
    labs(title = "position vs time - single id")
```
For each pair of load cycles after the first one there are a few points, 
then a long gap before the rest of the test.
This usually indicates that the cycle seperation code has included some of the
previous cycle, which can happen when the lower "peak" is wide and a bit noisy.

Fix that by finding large time gaps in the data and dropping points before the gap.

```{r}
#| label: drop_extra_points_start
# loading was already grouped by id, arm, plate, marker
# Calculate the time gap between points, 
# find gaps that are bigger than 2  the mean gap (to allow for a little jitter)
# get row of the first gap, or 1 if which() returns an empty vector (none found)
# and keep only points after that row to the end of the cycle
loading <- loading |> 
  group_by(id, arm, plate, marker, cycle) |>
  mutate(gap = time - lag(time), 
         big_gap = gap > 2 * mean(gap, na.rm = TRUE), # gap 0 is NA as no prior
         gap_pos = max(1, which(big_gap))) |>
  slice(max(gap_pos):n()) |> # could also use first(gap_pos), to get single value
  select(-c(gap, big_gap, gap_pos)) # Clean up the temporary columns
```

Plot again to check if it has worked

```{r}
#| label: test_cleaned

loading |>
#  filter(u < 0) |>
    ggplot() + 
    aes(x = time, y = u, colour = factor(cycle), group = uid) +
    geom_line() +
    facet_grid(plate + marker ~ id, scales = "free") +
    guides(colour=FALSE) +
    labs(title = "position vs time - by plate, marker and id")
```

## clean the motion data

There is some jiggle at the start of some cycles, 
where the construct does not always return to the same position.
Deal with that by creating incremental displacements for each cycle.

```{r}
loading <- loading |>
  mutate(across(c(u,v,w), \(m) m - first(m), .names = "inc_{.col}"))
```

Calculate total motion for each marker.
The `u` displacement is down (-ve), as is the `v` displacement generally.
The total absolute displacement will always be positive:
$$ motion = \sqrt{u^2 + v^2 + w^2} $$

```{r}
loading <- loading |> 
  mutate(motion = sqrt(inc_u^2 + inc_v^2 + inc_w^2)) 

loading |>
    ggplot() + 
    aes(x = load, y = motion, colour = factor(cycle), group = uid) +
    geom_line() +
    facet_grid(plate + marker ~ id, scales = "free") +
    guides(colour=FALSE) +
    labs(title = "motion vs load - by plate, marker and id", 
         x = "cycle", y = "total motion")
```

There still appear to be some kinks at the start of the data for 
the distal marker short plate for `MSD575`, `MSD663` and `MSD664`.
Checking, it looks like random variation, ecept for three cycles of `MSD664`.

```{r}
loading |>
    filter(id %in% c("MSD575", "MSD663", "MSD664"), marker == "distal", plate == "S") |>
    ggplot() + 
    aes(x = load, y = motion, colour = factor(cycle), group = uid) +
    geom_line() + geom_point() +
    facet_grid(id ~ ., scales = "free") +
    #guides(colour=FALSE) +
    labs(title = "motion vs load - by plate, marker and id", 
         x = "cycle", y = "total motion")

```

```{r}
loading |>
    filter(id == "MSD664", marker == "distal", plate == "S") |>
    ggplot() + 
    aes(x = load, y = motion, colour = factor(cycle), group = uid) +
    geom_line() + geom_point() +
    facet_grid( ~ cycle, scales = "free") +
    #guides(colour=FALSE) +
    labs(title = "motion vs load - by plate, marker and id", 
         x = "cycle", y = "total motion")

```

## fusion site motion

```{r}
# Make a plot for each plate

plot_p <- loading |>
    ggplot() + 
    aes(y = motion, x = load, colour = cycle, group = uid) +
    geom_line() +
    facet_grid(rows = vars(plate), cols = vars(id)) +
    labs(title = "load vs motion")
print(plot_p)
ggsave(filename = here::here("data", "motion_vs_cycle.png"), plot = plot_p)
```
```{r}
# Make a plot for each repeat

plot_p <- loading |>
    filter(id == 1) |>
    ggplot() + 
    aes(y = motion, x = load, colour = cycle, group = uid) +
    geom_line() +
    facet_grid(rows = vars(plate), cols = vars(rep)) +
    labs(title = "load vs motion")
print(plot_p)
ggsave(filename = here::here("data", "motion_vs_rep.png"), plot = plot_p)
```


## analysis

To get the stiffness of each test we fit a linear model by least squares.

We can analyse total motion at the end of the "metacarpal" (`position`), 
or the total motion at the fusion site (`motion`).
The latter is what is more clinically relevant, and by measuring at
the fusion site we remove the effect of differing bending in the 3D printed
bone analogues.

**Note that repeat 3 shows more variation.**
**This was also evident in the Instron raw data,** 
**and may be due to some electrical interference.**

```{r}
#| label: fit_lines

loading_models <- loading |>
  group_by(plate, id, rep, cycle) |>
  summarise(
    position_model = list(
      lm(load ~ position, data = pick(everything()))
    ),
    motion_model = list(
      lm(load ~ motion, data = pick(everything()))
    )
  )

# Extract the slope components (estimate) of the models
stiffness <- loading_models |>
  mutate(across(ends_with("model"), ~ purrr::map(.x, broom::tidy), 
                .names = "{.col}_tidy")) |> 
  tidyr::unnest(ends_with("tidy"), names_sep = "_") |>
  filter(position_model_tidy_term != "(Intercept)") |> 
  select(plate, id, rep, cycle, ends_with("estimate")) |>
  rename_with(~ stringr::str_remove(.x, "_model_tidy_estimate"))

# Plot the stiffness per cycle as a check
stiffness |> 
  filter(id == 1) |>
  ggplot() +
  aes(x= cycle, y = motion, colour = plate) +
  geom_point() +
  facet_grid(rows = vars(plate), cols = vars(rep)) +
  labs(title = "stiffness - plate 1 all repeats")

stiffness |> 
  filter(rep == 1) |> 
  ggplot() +
  aes(x= cycle, y = motion, colour = plate) +
  geom_point() +
  facet_grid(cols = vars(id), rows = vars(plate)) +
  labs(title = "stiffness - repeat 1 of each plate")

```

## check for effect of cycling

We have six plates.
Plate 1 of each type was tested multiple times, 
to measure the effect of repeat testing.
Plates 2 -- 6 of each type were each tested once.
Every test was made up of 60 cycles, with two cycles recorded every 10 - so
cycles 9, 10, 19, 20, 29, 30, 39, 40, 49, 50, 50, 60.

To test if repeated cycling is having an effect use a repeated 
measures ANOVA, as the cycles are repeated on the same subject for each test.

```{r}
#| label: repeated_measures_anova

# For repeats of the first plate
stiffness |> 
  ungroup() |> 
  filter(id == 1) |> 
  nest_by(plate) |> 
  mutate(model = list(aov(motion ~ cycle + Error(factor(rep)), data = data)),
         summary = list(broom::tidy(model))
         ) |> 
  tidyr::unnest(summary) |> 
  select(-data, -model)

# Across plates
stiffness |> 
  ungroup() |> 
  filter(rep == 1) |> 
  nest_by(plate) |> 
  mutate(model = list(aov(motion ~ cycle + Error(factor(id)), data = data)),
         summary = list(broom::tidy(model))
         ) |> 
  tidyr::unnest(summary) |> 
  select(-data, -model)

```
None of the tests are significant, so we can conclude that the stiffness is not 
changed by multiple cycles.
This means they can be averaged to reduce the variation between cycles.

Use the averaged stiffnesses in an ANOVA to look for differences due to plate type.
```{r}
#| label: compare_plates
#| warning: false
#| message: false

model <- stiffness |>
  summarise(stiffness = mean(motion), .groups = "keep") |>
  ungroup() |>
  filter(rep == 1) |>
  aov(stiffness ~ plate, data = _)

summary(model)
```

This shows no significant difference between the two plate designs.
The box plot shows this graphically.

```{r}
#| label: compare_plate_types
#| message: false
#| warning: false

# Compare average stiffness between plates

# This is for insertion into Matt's presentation, so want to use the font 
# Calibri, like the rest of the slides. These need to be installed previously.

extrafont::loadfonts() # makes other fonts available to R

plot_box <- loading |>
    filter(rep == 1) |>
    mutate(stiffness = mean(motion)) |>
    ggplot() + 
    aes(y = motion, x = plate) +
    geom_boxplot() +
    theme_bw(base_size = 14, base_family = "Calibri") 
    #labs(title = "plate stiffness")
print(plot_box)

ggsave(filename = here::here("data", "stiffness_between_plates.png"), 
       plot = plot_box, 
       width = 21, height = 12, units = "cm")
```

