---
title: "Medartis Wrist Plates"
subtitle: "Experiment 2 - bending and shear matched"
date: 2024-07-23
format: 
  html: default
  pdf: default
  docx: default
---

## setup

Load any needed libraries and define some constants.

```{r}
#| label: setup
#| warning: false

library(dplyr) # for manipulating dataframes
library(ggplot2) # for plotting

# If you don't have bluer install it with 
# inatall.packages("devtools")
# devtools::install_github(repo = "yadbor/bluer")
library(bluer) # routines for analysing mechanical test data


data_root <- here::here("data-raw", "Experiment 2", "Results", "VIC-3D data")

data_col_names <- "File Number,U [mm],V [mm],W [mm],Time_1,position,load"
data_cols <- stringr::str_split_1(data_col_names, ",")

```

## read the data files

Get all the `.csv` file names under the `data_root` folder 
and store them in a `list`.

Then run the list of file paths names through a pipeline 
that does the following:

1.  Name every list element with the `basename` from the path \
(i.e. just the file name with no path or extension)
1.  Read each .csv file into the list as a dataframe
1.  Keep only the wanted columns in each dataframe
1.  Stack the read dataframes, adding the names from the list as a column

```{r}
#| label: load_data_files
#| warning: false

# Find all .csv files under the data_root
file_path <- list.files(path = data_root, pattern = "*.csv", 
                         full.names = TRUE, recursive = TRUE)

all_results <- file_path |>        
  setNames(basename(file_path)) |> # name each item in the list with its basename
  purrr::map( \(df) readr::read_csv(df, show_col_types = FALSE) |> 
                    select(data_cols)
              ) |>
  bind_rows(.id = "filename")      # bind together, adding a column for the name

```

`all_results` is now a big dataframe 
(actually a [tibble](https://tibble.tidyverse.org/)) 
holding all the results from all tests, labelled with their file name.
This will work for studies like this one where all the meta-data are in the file names. 
For some studies some of the meta-data can be in the folder name as well. 
In those cases the final folder name can be extracted using `basename(pathname(full_path))`.

# extract test information

The filenames have the plate type, id and repeat encoded as:

"{Long|Short} Plate {id} T{repeat}_Data.csv"

Extract these into separate columns to identify each test.
Because the filenames don't have clean delimiters use a regex to break them up.

Some names have special issues, like the two that were repeated (T1.2 & T2.2).
Clean those up here as well.

```{r}
#| label: parse_filenames

all_results <- all_results |>
  tidyr::separate_wider_regex(cols = filename,
                              patterns = c(plate = "^\\w+", "\\D*", 
                                           id = "\\d+", "\\s*T", 
                                           rep = "[\\d\\.]+", ".*"),
                              cols_remove = FALSE
                              ) |>
  # Deal with special cases. 
  # In this study, T1.2 and T2.2 replace T1 and T2
  mutate(rep = stringr::str_remove(rep, "\\..*")) # delete everything after "."

```

Replace the column names with names that are easier to manipulate in code but 
still convey their purpose

```{r}
#| label: rename_columns

all_results <- all_results |>
  rename(frame = `File Number`, 
         u = `U [mm]`, v = `V [mm]`, w = `W [mm]`,
         time = `Time_1`)
```

## save clean data

Write the cleaned data to a `.csv` file before starting the analysis.

```{r}
#| label: save_data

all_results |> 
  readr::write_csv(file = here::here("data", "all_results.csv"))
```

## analyse cycles

Each test has multiple load-unload cycles. 
Data are recorded on the 9^th^ and 10^th^ cycles, and this is repeated 
every 10 cycles up to a total of 60 cyles.
We are interested in the loading part of each cycle.

Divide test into cycles by finding peaks and troughs.
The loading phase is the one going towards a peak, and the unloading phase is
towards a trough.

The `bluer` package has routines for finding peaks and labelling cycles 
and phases.

But first, as these tests are all in compression, the load and position are all 
negative.
Invert them both so that graphs read better for most people.

```{r}
#| label: label_cycles

cycles <- all_results |> 
  mutate(position = -1.0 * position, load = -1.0 * load) |> # Invert these axes
  group_by(plate, id, rep) |> # group each test
  mutate(as_tibble(bluer::label_cycles(position))) |> # and label the cycles
  mutate(cycle = factor(cycle)) |> # Make the cycle a factor for easier plotting
  mutate(uid = paste0(plate,id,rep,cycle)) # uniquely identify each cycle

# Check what we have done
cycles |> ggplot() + 
  aes(x = time, y = position, colour = seg, group = rep) +
  geom_point(aes(shape = rep)) +
  facet_grid(rows = vars(plate), cols = vars(id)) |>
  labs(title = "everything, everywhere, all at once")

# that plot was crowded, so do separate ones for each plate
#

cycles |> 
    filter(plate == "Long") |>
    ggplot() + 
    aes(x = time, y = position, colour = seg, group = rep) +
    geom_point() +
    facet_grid(rows = vars(id), cols = vars(rep)) +
    labs(title = "Long")

cycles |> 
    filter(plate == "Short") |>
    ggplot() + 
    aes(x = time, y = position, colour = seg, group = rep) +
    geom_point() +
    facet_grid(rows = vars(id), cols = vars(rep)) +
    labs(title = "Short")

```

Plot just the loading portion of each cycle, as load vs position.
```{r}
#| label: plot_load_position

loading <- cycles |> 
    filter(seg == "load")

loading |>
    filter(plate == "Short") |>
    ggplot() + 
    aes(x = position, y = load, colour = cycle, group = uid) +
    geom_line() +
    facet_grid(rows = vars(id), cols = vars(rep)) +
    labs(title = "load vs position - Short")

loading |>
    filter(plate == "Long") |>
    ggplot() + 
    aes(x = position, y = load, colour = cycle, group = uid) +
    geom_line() +
    facet_grid(rows = vars(id), cols = vars(rep)) +
    labs(title = "load vs position - Long")
```

## analysis

To get the stiffness of each test we fit a linear model by least squares.
```{r}
#| label: fit_lines

loading_models <- loading |>
  #group_by(uid) |>
  group_by(plate, id, rep, cycle) |>
  summarise(
    model = list(
      lm(load ~ position, data = pick(everything()))
    )
  )

loading_models |>   
  mutate(glance = purrr::map(model, broom::glance))|> 
  tidyr::unnest(glance)

# Extract the slope components of the models
lm_fits <- loading_models |> 
  mutate(tidy = purrr::map(model, broom::tidy)) |> 
  tidyr::unnest(tidy) |> 
  filter(term != '(Intercept)')

lm_fits

lm_fits |> ggplot() +
  aes(x= cycle, y = estimate, colour = plate) +
  geom_point() +
  facet_grid(rows = vars(id), cols = vars(rep))

lm_fits |> filter(rep == 1) |> ggplot() +
  aes(x= cycle, y = estimate, colour = plate) +
  geom_point() +
  facet_grid(cols = vars(id), rows = vars(plate))
```

## check for effect of cycling

We have six plates.
Plate 1 was tested multiple times, to measure the effect of repeat testing.
Plates 2 -- 6 were each tested once.
Every test was made up of 60 cycles, with two cycles recorded every 10 - so
cycles 9, 10, 19, 20, 29, 30, 39, 40, 49, 50, 50, 60.

To test if the number of cycles is having an effect use a repeated 
measures ANOVA, as the cycles are repeated on the same subject for each test.

```{r}
#| label: repeated_measures_anova

plate_1 <- lm_fits |>
  ungroup() |>
  filter(id == 1)
  # summarise(
  #   anova = list(aov(estimate ~ rep * cycle)))
  # )

# plate_1 |>
#   mutate(model = list(aov(formula = estimate ~ cycle + Error(rep/cycle), data = data)))

plate_1 |> 
  nest_by(plate) |>  
  mutate(model = list(aov(estimate ~ cycle + Error(rep/cycle), data = data))) |> 
  mutate(sum = list(broom::tidy(model))) |> 
  tidyr::unnest(sum)

# Check for any trend in the stiffness measured over time by fitting a line to
# stiffness vs cycle and checking if the the slope of that line is 
# significantly different to zero
# Note that there are a lot of comparisons, so use a correction for this.
# Also note that p.adjust doesn't work if the dataframe is grouped (returns p)
plate_1 |> 
  mutate(cycle = as.numeric(cycle)) |> 
  nest_by(plate, rep) |> 
  summarise(model = list(lm(formula = estimate ~ cycle, data = data))) |> 
  mutate(tidy = purrr::map(model, broom::tidy)) |> 
  tidyr::unnest(tidy) |> 
  filter(term == "cycle") |>
  ungroup() |>
  mutate(p.adj = p.adjust(p.value, "holm"))

```
None of the slopes are significant, so can conclude that the stiffness is not 
changed by multiple cycles.
This means they can be averaged to reduce the variation between cycles.

Use the averaged slopes in an ANOVA to look for differences due to plate type.
```{r}

model <- lm_fits |>
  summarise(stiffness = mean(estimate)) |>
  ungroup() |>
  filter(rep == 1) |>
  aov(stiffness ~ plate, data = _)

summary(model)
```

There are anomalies in the motion data.
The first cycles begins at u == 0, by definition (load about 20 N), 
but subsequent cycles begin at variable positions depending on exactly where 
the cycle start was found.
Likewise, some tests show an anomaly at the peak motion, probably due again to
not quite catching the turning points.

Fix these by starting at the first point where u >= 0. 
At the top can just trim the last two data points to avoid the issue, 
but it doesn't have much effect on the slope calculation so leave it for now.

```{r}
loading <- loading |> 
  group_by(plate, id, rep, cycle) |> 
  filter(u > 0) 

loading |>
  summarise(load=min(load), pos=min(position)) |> print(n=Inf)
```


## fusion site motion
```{r}
# plot U & V vs load
# Rearrange to long form for easier plotting in ggplot
loading_long <- loading |> 
    tidyr::pivot_longer(cols = c(u, v), names_to = "axis", values_to = "motion")
# Make a plot for each plate tpye
for (p in c("Long", "Short")) {
    plot_p <- loading_long |>
        filter(plate == p) |>
        ggplot() + 
        aes(y = motion, x = load, colour = cycle, group = uid) +
        geom_line() +
        facet_grid(rows = vars(axis), cols = vars(rep)) +
        labs(title = paste0("load vs position - ", p))
    print(plot_p)
}
```


## plot combined motion

```{r}

for (p in c("Long", "Short")) {
  plot_p <- loading |> 
    mutate(motion = sqrt(u^2 + v^2 + w^2)) |>
    filter(plate == p) |>
    ggplot() + 
    aes(y = motion, x = load, colour = cycle, group = uid) +
    geom_line() +
    ylim(c(0,0.6)) +
    facet_grid(rows = vars(id), cols = vars(rep)) +
    labs(title = paste0("load vs position - ", p))
  
  print(plot_p)
}

```

## compare combined motion

```{r}
# Plate 1 was used to test the effect of multiple repeats, 
# but for the inter-plate comparison only use the first rep to match the others.
plates_rep <- loading |> 
    mutate(motion = sqrt(u^2 + v^2 + w^2)) |>
    filter(rep == 1)

# Repeat the stiffness analysis above, but for the combined fusion site motion.
motion_models <- loading |> 
  mutate(motion = sqrt(u^2 + v^2 + w^2)) |>
  group_by(plate, id, rep, cycle) |>
  summarise(
    model = list(
      lm(load ~ motion, data = pick(everything()))
    )
  ) |> 
  mutate(tidy = purrr::map(model, broom::tidy)) |> 
  tidyr::unnest(tidy) |> 
  filter(term != '(Intercept)')

# Check for an effect of cycling with a repeated measures ANOVA
motion_models |>
  filter(id == 1)|> # The first plate id has multiple reps
  ungroup() |>
  nest_by(plate) |>  
  mutate(model = list(aov(estimate ~ cycle + Error(rep/cycle), data = data))) |> 
  mutate(sum = list(broom::tidy(model))) |> 
  tidyr::unnest(sum)

motion_models |>
  filter(rep == 1)|> # The first rep has multiple plate id
  ungroup() |>
  nest_by(plate) |>  
  mutate(model = list(aov(estimate ~ cycle + Error(id/cycle), data = data))) |> 
  mutate(sum = list(broom::tidy(model))) |> 
  tidyr::unnest(sum)


```



## notes



```{r}
#| label: scratchpad

# Make a vector of the actual cycle numbers we recorded:
#   9, 10, 19, 20, 29, 30 , 39, 40, 49, 50 , 59, 60
actual_cycles <- 10 * rep(1:6, each = 2) - rep(1:0, 6)
```

