---
title: "Medartis Wrist Plates"
subtitle: "Experiment 2 - bending and shear matched"
date: last-modified
format: 
  html: default
  pdf: default
  docx: default
---

## setup

Load any needed libraries and define some constants.

```{r}
#| label: setup
#| warning: false
#| message: false

library(dplyr) # for manipulating dataframes
library(ggplot2) # for plotting

# If you don't have bluer install it with 
# install.packages("devtools")
# devtools::install_github(repo = "yadbor/bluer")
library(bluer) # routines for analysing mechanical test data

```

## read the data files

Get all the `.csv` file names under the `data_root` folder 
and store them in a `list`.

Then run the list of file paths names through a pipeline 
that does the following:

1.  Name every list element with the `basename` from the path \
(i.e. just the file name with no path or extension)
1.  Read each .csv file into the list as a dataframe
1.  Keep only the wanted columns in each dataframe
1.  Stack the read dataframes, adding the names from the list as a column

```{r}
#| label: load_data_files
#| warning: false

data_root <- here::here("data-raw", "Experiment 2", "Results", "VIC-3D data")
# Find all .csv files under the data_root
file_path <- list.files(path = data_root, pattern = "*.csv", 
                         full.names = TRUE, recursive = TRUE)

all_results <- file_path |>        
  setNames(basename(file_path)) |> # name each item in the list with its basename
  purrr::map( \(df) readr::read_csv(df, show_col_types = FALSE)) |>
  bind_rows(.id = "filename")      # bind together, adding a column for the name

```

`all_results` is now a big dataframe 
(actually a [tibble](https://tibble.tidyverse.org/)) 
holding all of the results from all of the tests, labelled with their file name.

This approach works well for studies like this one,
where all the meta-data are in the file names. 

Some studies have some or all of the meta-data in the folder name.
This is the case for Instron raw data files, where the files are just numbered
and stored in a folder named for the test.

In those cases the final folder name of a path can be extracted using
`basename(dirname(full_path))`.

Studies with the metadata in several parts of the path can either be split by
repeatedly taking off the final component, or the path can be analysed with
the usual string manipulation tools (see `stringr`).

# extract test information

The filenames have the plate type, id and repeat encoded as:

"{Long|Short} Plate {id} T{repeat}_Data.csv"

Extract these into separate columns to identify each test.
Because the filenames don't have clean delimiters use a [regex](https://search.r-project.org/R/refmans/base/html/regex.html) 
to break them up.

Some file names have special issues, like the two tests that were repeated 
(T1.2 & T2.2).
Clean those up here as well.

```{r}
#| label: parse_filenames

all_results <- all_results |>
  tidyr::separate_wider_regex(cols = filename,
                              patterns = c(plate = "^\\w+", "\\D*", 
                                           id = "\\d+", "\\s*T", 
                                           rep = "[\\d\\.]+", ".*"),
                              cols_remove = FALSE
                              ) |>
  # Deal with special cases. 
  # In this study, T1.2 and T2.2 replaced T1 and T2
  mutate(rep = stringr::str_remove(rep, "\\..*")) # delete everything after "."

```

Replace the column names with names that are easier to manipulate in code but 
still convey their meaning.

```{r}
#| label: rename_columns

all_results <- all_results |>
  rename(frame = `File Number`, 
         u = `U [mm]`, v = `V [mm]`, w = `W [mm]`,
         time = `Time_1`)
```

## save clean data

Write the cleaned data to a `.csv` file before starting the analysis.
```{r}
#| label: save_data

all_results |> 
  readr::write_csv(file = here::here("data", "all_results.csv"))
```

## analyse cycles

Each test has multiple load-unload cycles. 
Data are recorded on the 9^th^ and 10^th^ cycles, and this is repeated 
every 10 cycles up to a total of 60 cycles.
We are interested in the loading part of each cycle.

Divide test into cycles by finding peaks and troughs, which are defined as local
maxima and minima.
The loading phase is the one going towards a peak, and the unloading phase is
towards a trough.

The `bluer` package has routines for finding peaks and labelling cycles 
and phases.

But first, as these tests are all in compression, both the load and position 
are negative.
Invert them both so that graphs read better for most people.

```{r}
#| label: label_cycles

cycles <- all_results |> 
  mutate(position = -1.0 * position, load = -1.0 * load) |> # Invert these axes
  group_by(plate, id, rep) |> # group each test
  mutate(as_tibble(bluer::label_cycles(position))) |> # and label the cycles
  mutate(cycle = factor(cycle)) |> # Make the cycle a factor for easier plotting
  mutate(uid = paste0(plate,id,rep,cycle)) # uniquely identify each cycle
```

Save the labelled data.
```{r}
#| label: write_cleaned_data
readr::write_csv(cycles, file = here::here("data", "experiment2_cycles.csv"))
```

If the data files have already been cleaned up and combined, 
just read the processed data from `experiment2_cycles.csv`.

These have been extracted from the raw files, compilied into a single table,
split into cycles (and the load and unload segment of each cycle), and had
the study metadata parsed out of the file name.

```{r}
#| label: read_data
#| eval: false # change this to read the file
cycles <- readr::read_csv(file = here::here("data", "experiment2_cycles.csv"),
                               show_col_types = FALSE)
```


```{r}
# Check what we have done
cycles |> ggplot() + 
  aes(x = time, y = position, colour = seg, group = rep) +
  geom_point(aes(shape = rep)) +
  facet_grid(rows = vars(plate), cols = vars(id)) |>
  labs(title = "everything, everywhere, all at once")

# that plot was crowded, so do separate ones for each sub-study
# First, the repeats of plate 1 (within plate)
cycles |> 
    filter(id == 1) |>
    ggplot() + 
    aes(x = time, y = position, colour = seg, group = rep) +
    geom_point() +
    facet_grid(rows = vars(plate), cols = vars(rep)) +
    labs(title = "Repeat tests on plate 1")
# Then the first repeat for each plate (between plates)
cycles |> 
    filter(rep == 1) |>
    ggplot() + 
    aes(x = time, y = position, colour = seg, group = rep) +
    geom_point() +
    facet_grid(rows = vars(plate), cols = vars(id)) +
    labs(title = "Test 1 for each plate")

```

Plot just the loading portion of each cycle, as load vs position.
```{r}
#| label: plot_load_position

loading <- cycles |> 
    filter(seg == "load")

# First, the repeats of plate 1 (within plate)
loading |>
    filter(id == 1) |>
    ggplot() + 
    aes(x = position, y = load, colour = cycle, group = uid) +
    geom_line() +
    facet_grid(rows = vars(plate), cols = vars(rep)) +
    labs(title = "load vs position - plate 1 all repeats")

# Then the first repeat for each plate (between plates)
loading |>
    filter(rep == 1) |>
    ggplot() + 
    aes(x = position, y = load, colour = cycle, group = uid) +
    geom_line() +
    facet_grid(rows = vars(id), cols = vars(plate)) +
    labs(title = "load vs position - repeat 1 of each plate")

```

## clean the motion data

There are anomalies in the motion data.
The first cycles begins at `u == 0`, by definition (load about 20 N), 
but subsequent cycles begin at variable positions depending on exactly where 
the cycle start was found.
Likewise, some tests show an anomaly at the peak motion, probably due again to
not quite catching the turning points.

Fix these by starting at the first point where `u >= 0`. 
At the top can just trim the last two data points to avoid the issue, 
but it doesn't have much effect on the slope calculation so leave it for now.

Then calculate the total motion at the fusion site from the vector sum of the 
three individual motions `(u, v, w)`

```{r}
loading <- loading |> 
  group_by(plate, id, rep, cycle) |> 
  filter(u >= 0) |>
  mutate(motion = sqrt(u^2 + v^2 + w^2)) 
```

## fusion site motion

```{r}
# Make a plot for each plate

plot_p <- loading |>
    filter(rep == 1) |>
    ggplot() + 
    aes(y = motion, x = load, colour = cycle, group = uid) +
    geom_line() +
    facet_grid(rows = vars(plate), cols = vars(id)) +
    labs(title = "load vs motion")
print(plot_p)
ggsave(filename = here::here("data", "motion_vs_cycle.png"), plot = plot_p)
```
```{r}
# Make a plot for each repeat

plot_p <- loading |>
    filter(id == 1) |>
    ggplot() + 
    aes(y = motion, x = load, colour = cycle, group = uid) +
    geom_line() +
    facet_grid(rows = vars(plate), cols = vars(rep)) +
    labs(title = "load vs motion")
print(plot_p)
ggsave(filename = here::here("data", "motion_vs_rep.png"), plot = plot_p)
```

## analysis

To get the stiffness of each test we fit a linear model by least squares.

We can analyse total motion at the end of the "metacarpal" (`position`), 
or the total motion at the fusion site (`motion`).
The latter is what is more clinically relevant, and by measuring at
the fusion site we remove the effect of differing bending in the 3D printed
bone analogues.

**Note that repeat 3 shows more variation.**
**This was also evident in the Instron raw data,** 
**and may be due to some electrical interference.**

```{r}
#| label: fit_lines

loading_models <- loading |>
  group_by(plate, id, rep, cycle) |>
  summarise(
    position_model = list(
      lm(load ~ position, data = pick(everything()))
    ),
    motion_model = list(
      lm(load ~ motion, data = pick(everything()))
    )
  )

# Extract the slope components (estimate) of the models
stiffness <- loading_models |>
  mutate(across(ends_with("model"), ~ purrr::map(.x, broom::tidy), 
                .names = "{.col}_tidy")) |> 
  tidyr::unnest(ends_with("tidy"), names_sep = "_") |>
  filter(position_model_tidy_term != "(Intercept)") |> 
  select(plate, id, rep, cycle, ends_with("estimate")) |>
  rename_with(~ stringr::str_remove(.x, "_model_tidy_estimate"))

# Plot the stiffness per cycle as a check
stiffness |> 
  filter(id == 1) |>
  ggplot() +
  aes(x= cycle, y = motion, colour = plate) +
  geom_point() +
  facet_grid(rows = vars(plate), cols = vars(rep)) +
  labs(title = "stiffness - plate 1 all repeats")

stiffness |> 
  filter(rep == 1) |> 
  ggplot() +
  aes(x= cycle, y = motion, colour = plate) +
  geom_point() +
  facet_grid(cols = vars(id), rows = vars(plate)) +
  labs(title = "stiffness - repeat 1 of each plate")

```
## save the calculated stiffness results

Write out the calculated stifness results.
These can be read to apply different analysis, without re-calculating from the raw data.

```{r}
#| label: save_stiffness
readr::write_csv(stiffness, file = here::here("data", "experiment2_stiffness.csv"))
```

## check for effect of cycling

We have six plates.
Plate 1 of each type was tested multiple times, 
to measure the effect of repeat testing.
Plates 2 -- 6 of each type were each tested once.
Every test was made up of 60 cycles, with two cycles recorded every 10 - so
cycles 9, 10, 19, 20, 29, 30, 39, 40, 49, 50, 50, 60.

To test if repeated cycling is having an effect use a repeated 
measures ANOVA, as the cycles are repeated on the same subject for each test.

```{r}
#| label: repeated_measures_anova

# For repeats of the first plate
stiffness |> 
  ungroup() |> 
  filter(id == 1) |> 
  nest_by(plate) |> 
  mutate(model = list(aov(motion ~ cycle + Error(factor(rep)), data = data)),
         summary = list(broom::tidy(model))
         ) |> 
  tidyr::unnest(summary) |> 
  select(-data, -model)

# Across plates
stiffness |> 
  ungroup() |> 
  filter(rep == 1) |> 
  nest_by(plate) |> 
  mutate(model = list(aov(motion ~ cycle + Error(factor(id)), data = data)),
         summary = list(broom::tidy(model))
         ) |> 
  tidyr::unnest(summary) |> 
  select(-data, -model)

```
None of the tests are significant, so we can conclude that the stiffness is not 
changed by multiple cycles.
This means they can be averaged to reduce the variation between cycles.

Use the averaged stiffnesses in an ANOVA to look for differences due to plate type.
```{r}
#| label: compare_plates
#| warning: false
#| message: false

model <- stiffness |>
  summarise(stiffness = mean(motion), .groups = "keep") |>
  ungroup() |>
  filter(rep == 1) |>
  aov(stiffness ~ plate, data = _)

summary(model)
```

This shows no significant difference between the two plate designs.

```{r}
#| label: mean differences
#| digits: 3
loading |>
  filter(rep == 1) |>
  mutate(stiffness = mean(motion)) |> 
  group_by(plate) |> 
  summarise(median = median(motion), mean = mean(motion)) |> 
  mutate(across(is.numeric, ~ round(.x, digits = 3)))
```

The box plot shows this graphically.

```{r}
#| label: compare_plate_types
#| message: false
#| warning: false

# Compare average stiffness between plates

# This is for insertion into Matt's presentation, so want to use the font 
# Calibri, like the rest of the slides. These need to be installed previously.

extrafont::loadfonts() # makes other fonts available to R

plot_box <- loading |>
    filter(rep == 1) |>
    mutate(stiffness = mean(motion)) |>
    ggplot() + 
    aes(y = motion, x = plate) +
    geom_boxplot() +
    theme_bw(base_size = 14, base_family = "Calibri") 
    #labs(title = "plate stiffness")
print(plot_box)

ggsave(filename = here::here("data", "stiffness_between_plates.png"), 
       plot = plot_box, 
       width = 21, height = 12, units = "cm")
```

